Alex's Notes and Text Fragments for the JSS Manuscript

### 2021-07-20 paper/paper.Rmd
emphasize the need to check for the prediction purpose of the model throughout the paper

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich Schiller University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena \AND'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com).
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatiotemporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST, blockCV or sperrorest.
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used.

  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and generic interface.
  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
  This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline, making it easier for users to apply a variety of different spatiotemporal resampling methods.
keywords:
  formatted: [cross-validation, predictive performance, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
  plain:     [cross-validation, predictive performance, autocorrelation, spatial, temporal, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{adjustbox}
  \usepackage{multirow}
  \usepackage{tabularx}
  \usepackage{placeins}
  \usepackage{scrextend}
  \usepackage{tablefootnote}
  \newcommand{\specialcell}[2][l]{%
    \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
tables: true
bibliography: paper.bib
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
---

@@

This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
The purpose of training a model on this sample is to make predictions on other, new fields within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
In this situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e. using other grid cells from the same fields, as if we wanted to predict within a field.
The results from this performance assessment would be over-optimistic, and perhaps badly so.
To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
If the model was to be applied to adjacent agricultural districts, i.e. outside the learning sample's spatial domain (*extra-domain* prediction; @brenning2005), it would even seem necessary to resample at a higher level of spatial aggregation, i.e. at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
The cross-validation resampling needed therefore depends as much on the prediction task as on the data structure or dependency at hand.

While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
\pkg{mlr3spatiotempcv} is such a toolbox.

@@

When $n_z$ is much larger than the desired number of folds, $k$, then a partitioning can be applied to the zones themselves. In this case, the zone indices $1, \ldots, n_z$ are grouped into $k$ equally sized subsets $\mathcal{I}_1, \ldots, \mathcal{I}_k$.
This approach has been applied, for example, in spatial CV at the agricultural field level [@pena2015].
In the following we will conceptually distinguish between CV *at the group (or block) level*, referring to the latter scenario, from spatial *block (or grouped)* CV, where the blocks or groups themselves define the CV partitions.

@@

Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method, this should not imply that this method is the best or only method suitable for this example task.
Even though this method is frequently used, it was mainly chosen because of it's clear visual grouping differences when comparing spatial to random partitioning.

In fact, most often multiple spatial partitioning methods can be used for a dataset.
It is recommended (required) that one familiarizes themselves with multiple methods and decide which method to choose based on the specific characteristics of the dataset at hand.
There are various variables influencing the partitioning method choice for an environmental dataset:

- Do natural groupings exists in the data?
- How dense are the observations scattered both in space and time?
- Are the predictor variables known to be highly affected by spatiotemporal autocorrelation by nature?

Based on these criteria users must choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by using all observations and eventually ignoring natural grouping patterns).
For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the strengths and weaknesses of the respective approach (either linked in the help file of the package or its respective upstream packages), hence we are refraining from repeating them again in this study.

In the example above, a cross-validation without hyperparameter tuning was shown.
In ML practice most often a nested CV should be desired to optimize model hyperparameter within the CV.
In such scenarios it is recommended to use the same spatial partitioning method for the inner loop (= tuning level) that was chosen for the outer one.
See @schratz2019 for more details as well as chapter 11 of [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-cv.html) [@lovelace2019].

@@

<!-- high autocor on the border of test sets -->
<!-- sptcv LLTO: discards many observations -->
All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is relatively high (unless the dataset inherits natural gaps between train and test sets).
Only methods that omit some observations (`"spcv_buffer"`, `"spcv_tiles"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce the impact of this issue.
The downside is the removal of observations from the dataset which possibly reduces the generalization power of the trained model and by this, depending on the specific observations that were removed, adds substantial bias to the CV results.

@@

<!-- missing: Guidelines how to choose a resampling method -->
The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods in R.
While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainty which method should be used for which dataset.
The answer to this question is not trivial and while there will most likely never be an easy answer to this, this topic needs to be discussed in more detail so that users feel more certain about their actions.
One option could be the proposal of a dataset characteristic checklist showing which characteristics make a certain dataset more suitable for specific methods.
This could then help users to make an appropriate and objective decision which resampling method might be a good fit for their dataset.

@@

<!-- signoff -->
Spatiotemporal cross-validation is still not yet a paradigm that is completely established in scientific workflows even though it is discussed intensively for more than a decade now.
Making most available methods easily accessible to users is an important step to foster the acceptance of such in the community and lead to less biased cross-validation results in environmental and ecological studies.

>>>>

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich Schiller University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena \AND'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com).
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatiotemporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST, blockCV and sperrorest.

  The goal of [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is to gather the available spatiotemporal resampling methods in R and make them available to users through a simple and generic interface.
  The latter is made possible by integrating the package directly into the [mlr3](https://mlr3.mlr-org.com) machine learning framework which already has support for generic non-spatiotemporal resampling methods such as random partitioning.
  This simplifies the step of integrating package specific syntax into an overarching machine learning pipeline, making it easier for users to apply a variety of different spatiotemporal resampling methods.
  The mlr3spatiotempcv package refrains from making suggestions which resampling methods users should apply as this decision mainly depends on the prediction purpose of the fitted model.
  No best method exists; only valid methods for the modeling purpose at hand.
keywords:
  formatted: [cross-validation, predictive performance, machine learning, autocorrelation, spatial, temporal, "\\proglang{R}"]
  plain:     [cross-validation, predictive performance, autocorrelation, spatial, temporal, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{adjustbox}
  \usepackage{multirow}
  \usepackage{tabularx}
  \usepackage{placeins}
  \usepackage{scrextend}
  \usepackage{tablefootnote}
  \newcommand{\specialcell}[2][l]{%
    \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
tables: true
bibliography: paper.bib
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
---

@@

This cross-validation bias is well-known in spatial as well as non-spatial prediction [@brenning2005; @brenning2008; @arlot2010; @roberts2017] and in forecasting [@bergmeir2018].
It is most easily understood from a predictive modelling perspective by focusing on the question where (and when) the model should be used for prediction.
In crop classification from remotely-sensed data, for instance, learning samples routinely contain multiple grid cells from a sample of fields with known crop type, for instance 2000 grid cells from 100 fields scattered across a large study region.
The purpose of training a model on this particular sample is to make predictions on other, new fields within the same geographic domain (*intra-domain* prediction; @brenning2005) --- not *within* the same field, which obviously presents only one crop type.
In this specific situation it would therefore seem rather unwise to train a model on a simple random subsample of grid cells, and test it on the remaining data, i.e. using other grid cells from the same fields, as if we wanted to predict within a field.
The results from this performance assessment would be over-optimistic, and perhaps badly so.
To mimic the predictive situation for which the model is trained, one would rather have to resample at the level of fields, not grid cells [@pena2015].
If the model was to be applied to adjacent agricultural districts, i.e. outside the learning sample's spatial domain (*extra-domain* prediction; @brenning2005), it would even seem necessary to resample at a higher level of spatial aggregation, i.e. at the level of agricultural districts within the learning sample, in order to realistically mimic the actual prediction task.
The cross-validation resampling needed therefore depends as much on the prediction task as on the data structure or dependency at hand.

While it is not the purpose of this article to recommend specific resampling schemes for specific use cases, this example may suffice to motivate the use of appropriate spatial and spatiotemporal cross-validation techniques, and the need for a unified framework and computational toolbox that accommodate a variety of prediction tasks that may be applicable to a broad range of application scenarios.
\pkg{mlr3spatiotempcv} is such a toolbox.

@@

When $n_z$ is much larger than the desired number of folds, $k$, then a partitioning can be applied to the zones themselves. In this case, the zone indices $1, \ldots, n_z$ are grouped into $k$ equally sized subsets $\mathcal{I}_1, \ldots, \mathcal{I}_k$.
This approach has been applied, for example, in spatial CV at the agricultural field level [@pena2015].
We would like to emphasize the conceptual distinction between CV *at the group (or block) level*, referring to the latter scenario, from *custom (or grouped)* CV, where the blocks or groups themselves define the CV partitions.

@@

Although the case study example in section \@ref(sec:case-study) used the `"spcv_coords"` method, this should not imply that this method is the best or only method suitable for this example task.
Even though this method is frequently used, it was mainly chosen because of it's clear visual grouping differences when comparing spatial to random partitioning.
Another advantage is its ability to adapt to the irregularly shaped study area of this example.
Last, there are no obvious spatial groupings to our knowledge which need to be accounted for, also with respect to prediction purposes of the fitted model.

Most often multiple spatial partitioning methods can be used for a dataset.
It is recommended (required) that one familiarizes themselves with multiple methods and decide which method to choose based on the specific characteristics of the dataset at hand and the prediction purpose of the fitted model, i.e. for which areas/zones a model should make predictions for.
There are various variables influencing the partitioning method choice for an environmental dataset:

- Do natural groupings exists in the data?
- How dense are the observations scattered both in space and time?
- Are existing predictor variables known to be highly affected by spatiotemporal autocorrelation by nature?

Based on these criteria users should choose a matching partitioning method that is either more restrictive (by discarding observations for fold creation) or more liberal (by using all observations and eventually ignoring natural grouping patterns).
For almost all methods implemented in \pkg{mlr3spatiotempcv}, there is a scientific publication describing the application areas of the respective approach (either linked in the help file of the package or its respective upstream packages), hence we are refraining from repeating them again in this study.

Last, a note on best practices for the example shown in \@ref(sec:case-study):
In this example a cross-validation without hyperparameter tuning was applied.
In ML practice most often a nested CV should be desired to optimize model hyperparameter within the CV.
In such scenarios it is recommended to use the same spatial partitioning method for the inner loop (= tuning level) that was chosen for the outer one for consistency reasons.
See @schratz2019 for more details as well as chapter 11 of [Geocomputation with R](https://geocompr.robinlovelace.net/spatial-cv.html) [@lovelace2019].

@@

Even though the prediction purpose should be the first decision criteria for choosing an appropriate spatiotemporal resampling method, we would like to discuss the individual methods in more detail in the following section.

<!-- high autocor on the border of test sets -->
<!-- sptcv LLTO: discards many observations -->
All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is relatively high (unless the dataset inherits natural groupings or gaps between train and test sets).
Methods that omit some observations (`"spcv_buffer"`, `"spcv_tiles"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce the impact of autocorrelation in scenarios without a clear grouping or sufficient distance between train and test set observations.
The downside is the removal of observations from the dataset which possibly reduces the generalization power of the trained model and by this, depending on the specific observations that were removed, adds substantial bias to the CV results.

@@

<!-- missing: Guidelines how to choose a resampling method -->
The \pkg{mlr3spatiotempcv} package is the first package to bundle and categorize spatiotemporal resampling methods in R.
While various spatiotemporal resampling methods exists nowadays to account for spatiotemporal autocorrelation in modeling, there is still lots of uncertainty which method should be used for which dataset.
The answer to this question is not trivial and while there will most likely never be an easy answer to this, users are advised to first think about the prediction purpose of the model and then about reducing autocorrelation between train and test sets to reduce the impact of spatiotemporal autocorrelation.
One option could be the proposal of a dataset characteristic checklist showing which characteristics make a certain dataset more suitable for specific methods.
This could then help users to make an appropriate and objective decision which resampling method might be a good fit for their dataset and prospective prediction task.

@@

<!-- signoff -->
Spatiotemporal cross-validation is still not yet a paradigm that is completely established in scientific workflows even though it is discussed intensively for more than a decade now.
Making most available methods easily accessible to users is an important step to foster the acceptance of such in the community and to lead to less biased cross-validation results in environmental and ecological studies.
